Cycle Route Uptake and Scenario Estimation (CRUSE): an approach for developing strategic cycle network planning tools

### 2024-06-23 paper.tex
R2 version

\setstretch{2}
\newpage{}

@@

Tools for estimating geographical distribution of cycling potential have
advanced substantially in recent years. Developed by researchers and
developers in academic, public and private sectors, they have evolved
from simple area-based static models tailored to specific regions to
more complex and (in some cases) more generalizable tools. These tools
can be considered as a specific application of the concept of planning
support systems (PSS) , which are designed to help planners and
decision-makers to make more informed decisions \citep{geertman2009}.
Outputs include maps with results available at several levels of
analysis with levels of availability ranging from being only available
to researchers and in static maps to open access mapping systems
available to the public.

@@

\\
\toprule
Approach & Coverage and input data & Outputs and availability & License\\
\midrule
\cellcolor{gray!10}{Bicycle share model, Parkin et al. (2007)} & \cellcolor{gray!10}{England and Wales, Journeys to work OD census data, at the small-area (wards) level} & \cellcolor{gray!10}{Static tables in academic paper} & \cellcolor{gray!10}{Proprietary}\\
Analysis of  Cycling Potential (ACP), Transport for London (2010, 2016) & London, London Travel Demand Survey & Static tables, graphs, and heatmaps, not publicly available & Proprietary\\
\cellcolor{gray!10}{Prioritization index, Larsen et al. (2013)} & \cellcolor{gray!10}{Montreal, Survey, Road safety, and OD data} & \cellcolor{gray!10}{Map-based heatmap, not publicly available} & \cellcolor{gray!10}{Proprietary}\\
Usage intensity index, Zhang et al. (2014) & Belo Horizonte, Survey, census, and OD data & Static tables and maps, not publicly available & Proprietary\\
\cellcolor{gray!10}{The Cycling Potential Tool (CPT), Phillips and Range (2017)} & \cellcolor{gray!10}{Scotland, Environmental and socioeconomic data, at the small area (output areas) level} & \cellcolor{gray!10}{Maps showing cycling potential in each area} & \cellcolor{gray!10}{Proprietary}\\
\addlinespace
Propensity to Cycle Tool (PCT), Lovelace et al. (2017) & England and Wales, Journeys to work and school OD census data, at the small area (LSOA) level & Online maps, graphs, tables, publicly available at www.pct.bike & Open source\\
\cellcolor{gray!10}{The    Gross Potential for Cycling tool (CPC), Silva et al. (2021 and 2022)} & \cellcolor{gray!10}{21 Portuguese cities, Land use and socio-demographic data, at the small area (census tract) level} & \cellcolor{gray!10}{Static maps showing cycling potential in different areas} & \cellcolor{gray!10}{NA}\\
\bottomrule

@@

Other tools that generate areal data to prioritize investment include a
study that generated a `prioritization index' for Montreal, Canada
\citep{larsen_build_2013}, a `usage intensity index' based on stated
preferences in Belo Horizonte, Brazil, and a Cycling Potential Tool
(CPT) for Scotland \citep{phillips_development_2017}. The CPT is
composed of the base environmental module --- based on eight weighted
factors (population density, hilliness, physical barriers, access to
services, existing cycling mode share, distance to work and school, and
road speed) --- and the quality of service module, which evaluates each
area of interest based on eight factors (surface condition, adjacent
cyclists, comfort factor, conflict, distance between junctions, slope,
access to services, and origin/destination).

A more recent areal-based approach is the designed the Gross Potential
for Cycling (GPC) tool to prioritize areas for cycling infrastructure
and other cycling measures \citep{silva_gross_2021}. The GPC uses two
sets of indicators: population-based indicators (age, potential demand
density, employment density and motorization rate) and area-based
indicators (accessibility, education, public transport, connectivity,
land use mix and relative performance). These indicators are calculated
considering topography, road hierarchy, and average congestion and
presented on a scale from 5 to 10. They are then combined into an
overall score, weighted by their impact on cycling according to the
literature. The ranking of areas by the GPC are
\href{https://boost.up.pt/en/ferramentas/gpc}{publicly available} and
its practical value has been assessed through workshops
\citep{silva_revealing_2022}.

Route-based tools emerged with increasing availability of
origin-destination data, routing engines that can assign trips to
networks, and improvements in computer hardware and software needed to
generate route networks. A prominent example is the the Propensity to
Cycle Tool (PCT), first developed to estimate current and future levels
of cycling at desire line, zone, route, and route network levels for
case study cities \citep{lovelace_mapping_2016}. The approach was
scaled-up to estimate estimate the potential benefits of uptake at zone
and desire line levels nationally, and launched as a publicly available
web application in 2017 \citep{lovelace2017}. Extensions of the PCT
approach have included estimation of benefits at the individual level
\citep{woodcock2018}, addition of travel to school network
\citep{goodman2019}, and improved modeling of impacts on health,
environmental and distributional outcomes \citep{woodcock2021}.
Initially developed just for England, the PCT was extended to cover all
of Wales (for commuter data only) in 2018.

@@

@@

\begingroup\fontsize{8}{10}\selectfont

\begin{longtable}[t]{>{\raggedright\arraybackslash}p{5em}>{\raggedright\arraybackslash}p{12em}>{\raggedright\arraybackslash}p{15em}>{\raggedright\arraybackslash}p{15em}}

@@

\toprule
Trip purpose & Description & Primary source(s) & Confidence\\
\midrule
\cellcolor{gray!10}{Work (commute)} & \cellcolor{gray!10}{Commuting to / from workplaces} & \cellcolor{gray!10}{Census 2016 POWSCAR Data} & \cellcolor{gray!10}{High confidence - data both origins and destinations for trip purposes}\\
Primary Education & Primary school trips & Census 2016 POWSCAR Data, Schools database & High confidence - data both origins and destinations for trip purposes\\
\cellcolor{gray!10}{Secondary Education} & \cellcolor{gray!10}{Secondary school trips} & \cellcolor{gray!10}{Census 2016 POWSCAR Data, Schools database} & \cellcolor{gray!10}{High confidence - data both origins and destinations for trip purposes}\\
Tertiary Education & Tertiary education trips & Census 2016 POWSCAR Data - Geodirectory categories & High confidence - data both origins and destinations for trip purposes\\
\cellcolor{gray!10}{Social} & \cellcolor{gray!10}{NHTS classification (e.g. restaurants, cinemas, gyms etc.)} & \cellcolor{gray!10}{Trip rates from NHTS, Geodirectory businesses classified according to relevant NACE codes} & \cellcolor{gray!10}{Medium confidence}\\
\addlinespace
Shopping & NHTS classificationfrom supermarkets and shops & Trip rates from NHTS, Geodirectory businesses classified according to relevant NACE codes & Medium confidence\\
\cellcolor{gray!10}{Personal / other} & \cellcolor{gray!10}{Combination of two NHTS classifications, and includes trips to medical, personal services and others} & \cellcolor{gray!10}{Trip rates from NHTS, Geodirectory businesses classified according to relevant NACE codes} & \cellcolor{gray!10}{Medium confidence}\\
Tourism / recreational & Trips by non-resident visitors to / from visitor attractions & Geodirectory listings for accommodation (origins) and destinations (destinations), Failte Ireland accommodation database, Failte Ireland attraction database & Low confidence - lack of data regarding tourist trip rates\\
\bottomrule

@@

\endgroup{}

@@

\includegraphics{fig-user-stories-1.pdf}

@@

@@

\subsection{Stakeholder engagement and feedback}\label{sec-qualitative}

@@

>>>>

\setstretch{2}
Note: preprint now available here: https://osf.io/preprints/osf/a72yh

Please cite as

Lovelace, Robin, Joey Talbot, Eugeni Vidal-Tortosa, Hussein Mahfouz,
Elaine Brick, Peter Wright, Gary O'Toole, et al.~2024. ``Cycle Route
Uptake and Scenario Estimation (CRUSE): An Approach for Developing
Strategic Cycle Network Planning Tools.'' OSF Preprints. March 29.
doi:10.31219/osf.io/a72yh.

This is an open and reproducible paper. See
\href{https://github.com/cruse-bike}{github.com/cruse-bike} for the code
and data used to generate the results presented here, and the code
underlying the web app hosted at \href{https://cruse.bike/}{cruse.bike}.

\newpage{}

@@

Tools for estimating geographical distribution of cycling potential have
advanced substantially in recent years. Developed by researchers and
developers in academic, public and private sectors, they have evolved
from simple area-based static models tailored to specific regions to
more complex and (in some cases) more generalizable tools. These tools
can be considered as a specific application of the concept of planning
support systems (PSS), which are designed to help planners and
decision-makers to make more informed decisions \citep{geertman2009}.
Outputs include maps with results available at several levels of
analysis with levels of availability ranging from being only available
to researchers and in static maps to open access mapping systems
available to the public.

@@

\\
\toprule
Approach & Coverage and input data & Outputs and availability & License\\
\midrule
\cellcolor{gray!10}{Bicycle share model, Parkin, J. et al. (2007)} & \cellcolor{gray!10}{England and Wales, Journeys to work OD census data, at the small-area (wards) level} & \cellcolor{gray!10}{Static tables in academic paper} & \cellcolor{gray!10}{Proprietary}\\
Analysis of Cycling Potential (ACP), Transport for London (2010 and 2016) & London, London Travel Demand Survey & Static tables, graphs, and heatmaps, not publicly available & Proprietary\\
\cellcolor{gray!10}{{}[Bikeability index/Bike Score]\{style="color: blue;"\}, [Winters, M. et al. (2013 and 2016)]\{style="color: blue;"\}} & \cellcolor{gray!10}{{}[Metro Vancouver region/160 US and Canadian cities]\{style="color: blue;"\}, [Cycling infrastructure, topography, destinations and road connectivity data]\{style="color: blue;"\}} & \cellcolor{gray!10}{{}[Map-based heatmap, not publicly available]\{style="color: blue;"\}} & \cellcolor{gray!10}{{}[Proprietary]\{style="color: blue;"\}}\\
Prioritization index, Larsen, J. et al. (2013) & Montreal, Survey, Road safety, and OD data & Map-based heatmap, not publicly available & Proprietary\\
\cellcolor{gray!10}{Usage intensity index, Zhang, D. et al. (2014)} & \cellcolor{gray!10}{Belo Horizonte, Survey, census, and OD data} & \cellcolor{gray!10}{Static tables and maps, not publicly available} & \cellcolor{gray!10}{Proprietary}\\
\addlinespace
The Cycling Potential Tool (CPT), Phillips, L., \& Range, A. (2017) & Scotland, Environmental and socioeconomic data, at the small area (output areas) level & Maps showing cycling potential in each area & Proprietary\\
\cellcolor{gray!10}{Propensity to Cycle Tool (PCT), Lovelace, R. et al. (2017)} & \cellcolor{gray!10}{England and Wales, Journeys to work and school OD census data, at the small area (LSOA) level} & \cellcolor{gray!10}{Online maps, graphs, tables, publicly available at www.pct.bike} & \cellcolor{gray!10}{Open source}\\
{}[Cyclist Routing Algorithm for Network Connectivity (CRANC)]\{style="color: blue;"\}, [Gehrke, S. R. et al. (2020)]\{style="color: blue;"\} & {}[Boston region]\{style="color: blue;"\}, [Road segment characteristics, elevation, and destination acessibility data, at the census blocks level]\{style="color: blue;"\} & {}[Graphs and heatmaps, not publicly available]\{style="color: blue;"\} & {}[NA]\{style="color: blue;"\}\\
\cellcolor{gray!10}{The Gross Potential for Cycling tool (CPC), Silva, C. et al. (2021 and 2022) [and Lopez et al. (2021)]\{style="color: blue;"\}} & \cellcolor{gray!10}{21 Portuguese cities, Land use and socio-demographic data, at the small area (census tract) level} & \cellcolor{gray!10}{Static maps showing cycling potential in different areas} & \cellcolor{gray!10}{NA}\\
{}[The relative equity impact of bicycle planning (TIRE)]\{style="color: blue;"\}, [Cunha, I., \& Silva, C. (2023)]\{style="color: blue;"\} & {}[Lisbon]\{style="color: blue;"\}, [Cycling infrastructure, topog­raphy, and location of points of interest data, at the hexagon cells level]\{style="color: blue;"\} & {}[Graphs and heatmaps, not publicly available]\{style="color: blue;"\} & {}[NA]\{style="color: blue;"\}\\
\bottomrule

@@

Other tools that generate areal data to prioritize investment include a
study that generated a {`bikeability index' for the Metro Vancouver
region \citep{winters_mapping_2013}, later renamed Bike Score and
available for over 160 US and Canadian cities \citep{winters_bike_2016}
-- its methodology is
\href{https://www.walkscore.com/bike-score-methodology.shtml}{openly
accessible}; a} `prioritization index' for Montreal, Canada
\citep{larsen_build_2013}; a `usage intensity index' based on stated
preferences in Belo Horizonte, Brazil \citep{zhang_prioritizing_2014};
and a Cycling Potential Tool (CPT) for Scotland
\citep{phillips_development_2017}. The CPT is composed of the base
environmental module --- based on eight weighted factors (population
density, hilliness, physical barriers, access to services, existing
cycling mode share, distance to work and school, and road speed) --- and
the quality of service module, which evaluates each area of interest
based on eight factors (surface condition, adjacent cyclists, comfort
factor, conflict, distance between junctions, slope, access to services,
and origin/destination).

A more recent areal-based approach is the Gross Potential for Cycling
(GPC) tool to prioritize areas for cycling infrastructure and other
cycling measures \citep{silva_gross_2021}. The GPC uses two sets of
indicators: population-based indicators (age, potential demand density,
employment density and motorization rate) and area-based indicators
(accessibility, education, public transport, connectivity, land use mix
and relative performance). These indicators are calculated considering
topography, road hierarchy, and average congestion and presented on a
scale from 5 to 10. They are then combined into an overall score,
weighted by their impact on cycling according to the literature. The
ranking of areas by the GPC are
\href{https://boost.up.pt/en/ferramentas/gpc}{publicly available} and
its practical value has been assessed through workshops
\citep{silva_revealing_2022}. {The GPC tool has also been used to
investigate the combination of factors contributing to high levels
cycling potential \citep{lopes_impact_2021}. Another recent areal-based
approach is the Relative Equity Impact of Bicycle Planning (TIRE)
\citep{cunha_assessing_2023} tool. The TIRE tool assesses the impact of
cycling network allocation on the accessibility levels of different
socioeconomic groups.}

Route-based tools emerged with increasing availability of
origin-destination data, routing engines that can assign trips to
networks, and improvements in computer hardware and software needed to
generate route networks. A prominent example is the the Propensity to
Cycle Tool (PCT), first developed to estimate current and future levels
of cycling at desire line, zone, route, and route network levels for
case study cities \citep{lovelace2016}. The approach was scaled-up to
estimate estimate the potential benefits of uptake at zone and desire
line levels nationally, and launched as a publicly available web
application in 2017 \citep{lovelace2017}. Extensions of the PCT approach
have included estimation of benefits at the individual level
\citep{woodcock2018}, addition of travel to school network
\citep{goodman2019}, and improved modeling of impacts on health,
environmental and distributional outcomes \citep{woodcock2021}.
Initially developed just for England, the PCT was extended to cover all
of Wales (for commuter data only) in 2018.

@@

{Another route-based tool is The Cycling Routing Algorithm for Network
Connectivity (CRANC) \citep{gehrke_cycling-focused_2020}. The CRANC tool
evaluates the benefits of new bicycle facilities for the accessibility
of various populations and neighbourhoods.}

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.0653}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2869}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.4460}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2017}}@{}}

@@

\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Trip purpose
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Description
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Primary source(s)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Confidence
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Work (commute) & Commuting to / from workplaces & Census 2016 POWSCAR
Data & High confidence - data both origins and destinations for trip
purposes \\
Primary Education & Primary school trips & Census 2016 POWSCAR Data,
Schools database & High confidence - data both origins and destinations
for trip purposes \\
Secondary Education & Secondary school trips & Census 2016 POWSCAR Data,
Schools database & High confidence - data both origins and destinations
for trip purposes \\
Tertiary Education & Tertiary education trips & Census 2016 POWSCAR Data
- Geodirectory categories & High confidence - data both origins and
destinations for trip purposes \\
Social & NHTS classification (e.g.~restaurants, cinemas, gyms etc.) &
Trip rates from NHTS, Geodirectory businesses classified according to
relevant NACE codes & Medium confidence \\
Shopping & NHTS classificationfrom supermarkets and shops & Trip rates
from NHTS, Geodirectory businesses classified according to relevant NACE
codes & Medium confidence \\
Personal / other & Combination of two NHTS classifications, and includes
trips to medical, personal services and others & Trip rates from NHTS,
Geodirectory businesses classified according to relevant NACE codes &
Medium confidence \\
Tourism / recreational & Trips by non-resident visitors to / from
visitor attractions & Geodirectory listings for accommodation (origins)
and destinations (destinations), Failte Ireland accommodation database,
Failte Ireland attraction database & Low confidence - lack of data
regarding tourist trip rates \\

@@

@@

\includegraphics[width=3.31in,height=\textheight]{images/fig5.png}

@@

\subsection{\texorpdfstring{{Stakeholder
engagement}}{Stakeholder engagement}}\label{sec-stakeholder-engagement}

{The CRUSE Tool was developed for, and in close collaboration with,
practitioners in Ireland. Although the overall approach of the project
was quantitative, stakeholder feedback and qualitative data was deemed
essential to ensure the tool was fit for purpose, mitigating against the
risk that the project would be `technocentric' and a `problem looking
for a solution'. From the outset of the project, we aimed to follow
\citet{coppola2022}`s advocacy for a ``human-centric approach in which
personalization, purpose, ethics and inclusivity are key'', rather than
an 'off-the-shelf' product that failed to understand, let alone meet,
local needs. As outlined by \citet{coppola2022}, the `digital
transformation' taking place in transport planning (and many other
sectors) has the potential to increase existing inequalities
\citep{coppola2022}. A wide range of methods and measure can be taken to
ensure that digital tools are accessible to their recipients.}

{The stakeholder engagement was carried out through workshops with
stakeholders in case study area, an approach that is documented with
reference to participatory development of digital tools for supporting
energy transitions \citep{hewitt2020}. We held a total of three
workshops over the course of the project and one open-ended live demo
during a 2-day event in Sligo in which we collected feedback from dozens
of people. Other than the live demo, all of the workshops took place
online.}

{The first two workshops took place in May 2022, during which the tool
was introduced to key staff involved in cycle network planning in
Kildare and Limerick counties. The county-specific approach allowed the
workshop to focus on specific areas and key corridors. The local
transport planners were asked to provide feedback in a semi-structured
interview approach, with questions on the evidence they currently have
access, how they may be able to use the tool, and requests for new
features. The third and final workshop took place on the
7\textsuperscript{th} December 2023 and included descriptions of recent
changes to the tool, with an aim being to get feedback on the new
recreational and cycle touring evidence, with representatives from
county-level transport planning bodies, Sport Ireland, and Waterways
Ireland.}

{The open-ended live demo was undertaken at a stall at TII's National
Roads and Greenways Conference in Sligo in September 2022, where much of
the feedback was verbal and not collected. Much of this feedback from
the workshops and live demo outlined above was in the form of ad-hoc
comments and suggestions and not all of the qualitative data received
was recorded in a structured way, a limitation that we refer back to in
Section~\ref{sec-discussion}.}

@@

\subsection{Stakeholder feedback}\label{sec-qualitative}

@@

{A final limitation of the approach relates to \emph{how} it was
developed. As described in Section~\ref{sec-stakeholder-engagement}, the
development of the tool was participatory, but only insofar as people
who attended our workshops could provide feedback. The people who did
provide feedback were deliberately chosen as the target audience. In
future work we would like to collect feedback from a wider range of
people, including campaigners, service providers (e.g.~staff involved in
travel planning for schools and new developments), and people who not
involved in the planning process.}
